Enhancing Adversarial Robustness for Deep Metric Learning

### 2021-10-29 robdml.tex
remove sotaplot
tex

To validate the effectivenss of HM, we start from using the hardness of another
benign triplet sampled with a different triplet sampler as $H_\mathsf{D}$, as
the existing triplet samplers are known not to result in model collapse in
regular training.

@@

Due to the flexibility of HM, any of the existing (and future) triplet samplers
can be used for calculating $H_\mathsf{S}$ and $H_\mathsf{D}$.
%
But not all possible combinations are expected to be effective or efficient in
terms of adversarial training, as the \emph{source} hardness of a notable
portion of the given triplets should be increased.
%
Thus, we first sort the existing samplers, namely Random,
Semihard~\cite{facenet}, Softhard~\cite{revisiting},
Distance-weighted~\cite{distance} (\emph{abbr}., Distance), and the
within-batch Hardest negative sampler, based on the approximated mean hardness
of their outputs, as shown in~\cref{tab:hsort}.
%
Then we adversarially train models on CUB dataset with all combinations
respectively, and summarize the R@1 and ERS in \cref{tab:desth}.

% 2. table: left and right

As shown in the last three columns of \cref{tab:desth}, almost all combinations
lead to model collapse as long as $H_\mathsf{D}$ is calculated from a Softhard,
Distance, or Hardest triplet.
%
Due to the small $E[H_\mathsf{D}-H_\mathsf{S}]$ in the combination of Distance
and Hardest triplets, the adversarial examples will be weak and the model does
collapse, but the robustness gain is negligible.
%
Although $E[H]$ of Softhard is less than that of Distance or Hardest,
some hard adversarial examples are still created due to its large
$\text{Var}[H]$, which still result in a slow collapse.
\todo{merge this para into upper triangular and lower triangular}

@@

For cases in the upper triangular of \cref{tab:desth} where $E[H_\mathsf{S}]
\leqslant E[H_\mathsf{D}]$, most given triplets will be turned adversarial with
an increased hardness.
%
Although most of such cases end up with model collapse, the preliminary result
suggests that $\text{HM}[\mathcal{R},\mathcal{M}]$ is still effective in
improving the robustness without model collapse, at an penalty on the R@1.

@@

For cases in lower triangular of \cref{tab:desth}, where $E[H_\mathsf{S}]
\geqslant E[H_\mathsf{D}]$, a large portion of given triplets will be
unchanged according to \cref{eq:hm}, and hence lead to weak robustness.
%
As an exception, $\text{HM}[\mathcal{S},\mathcal{M}]$ is still effective in
improving adversarial robustness, where a notable number of adversarial
examples are created due the high $\text{Var}[H]$ of Softhard.
%
%
%
In other words, HM in this case can be interpreted as mixing some unperturbed
softhard triplets and some semihard adversarial triplets.

@@

\begin{figure}[t]
	\includegraphics[width=\columnwidth]{sotaplot.pdf}
	\caption{sota plot}
\end{figure}

>>>>

As discussed in \cref{sec:31}, we start from the $H_\mathsf{D}$ calculated from
a harder benign triplet sampled in a different strategy, such as Random,
Semihard~\cite{facenet}, Softhard~\cite{revisiting},
Distance-weighted~\cite{distance} (\emph{abbr}., Distance), or the within-batch
Hardest negative sampling strategy, because we know these strategies do not
result in model collapse in regular training.

@@

HM is flexible so that any existing or future triplet sampling strategy
can be used for the source triplet or calculating $H_\mathsf{D}$.
%
But not all potential combinations are expected to be effective for HM, as it
will not create an adversarial triplet when $H_\mathcal{S}\geqslant
H_\mathcal{D}$.
%
Thus, we sort the strategies based on the mean hardness of their
outputs in~\cref{tab:hsort}.
%
Then we adversarially train models on CUB dataset with all combinations
respectively, and summarize their R@1 and ERS in \cref{tab:desth}.

@@

For cases in the upper triangular of \cref{tab:desth} where $E[H_\mathsf{S}]
\leqslant E[H_\mathsf{D}]$, most of the given triplets will be turned
adversarial.
%
Although almost all of them end up with model collapse, the
$\text{HM}[\mathcal{R},\mathcal{M}]$ is still effective in improving the
robustness, with an expected performance drop in R@1.
%
The combination of Distance and Hardest triplets does not trigger model
collapse due to the small $E[H_\mathsf{D}-H_\mathsf{S}]$, which leads to weak
adversarial examples and a negligible robustness gain.

For cases in lower triangular of \cref{tab:desth}, where $E[H_\mathsf{S}]
\geqslant E[H_\mathsf{D}]$, a large portion of given triplets will be unchanged
according to \cref{eq:hm}, and hence lead to weak robustness.
%
As an exception, $\text{HM}[\mathcal{S},\mathcal{M}]$ is still effective in
improving adversarial robustness, where a notable number of adversarial
examples are created due the high $\text{Var}[H]$ of Softhard.
%
Although $E[H]$ of Softhard is less than that of Distance or Hardest, some hard
adversarial examples are still created
%
\footnote{Differently, Softhard also samples a
hard positive instead of a random positive besides a hard negative.
%
As a result, the hardness of a small number of Softhard triplets will be 
greater than that of a given Hardest triplet.}
%
due to its large $\text{Var}[H]$, which
still result in a slow collapse.

In summary, HM creates mini-batches mixing some unperturbed source triplets and
some adversarial triplets.
%
The $\text{HM}[\mathcal{R}, \mathcal{M}]$ and $\text{HM}[\mathcal{S},
\mathcal{M}]$ achieve such balanced mixtures.
%
Empirically, the hardness range of Semihard strategy is appropriate for
$H_\mathcal{D}$.

@@

@@

\begin{comment}
\begin{figure}[t]
	\includegraphics[width=\columnwidth]{sotaplot.pdf}
	\caption{sota plot}
\end{figure}
\end{comment}
